\textbf{Real-time guitar transcription using\\Fourier transform based methods}\\A pitch estimation framework and overtone sieve algorithm

### 2021-06-18 paper.tex
2e versie bijna klaar (experimenten en conclusie over)

@@

\section*{Abstract}
In this paper, we explore real-time transcription of a electric guitar signal. We focus on Fourier transform based methods, as this transform can be performed very fast. Using the obtained frequency domain, we select significant peaks using a Gaussian envelope. These peaks contain many overtones, which are filtered using our overtone sieve algorithm.

@@

% benadrukken dat er weinig nieuws is, maar combinatie is van bestaande dingen
\section{Introduction}
Automatic Music Transcription (AMT) is a field of study which tries to convert acoustic recordings into some form of (digital) music notation~\cite{survey1}. This task can be divided into two subproblems: pitch estimation and onset detection~\cite{survey2}. Instead of trying to convert a song into music notation, this paper will focus on translating a guitar signal into a set of played notes in real-time. This problem is very similar to AMT, as both require pitch estimation and onset/offset detection, however, we also have take latency into account.

Single pitch estimation can be accurately performed using Fourier transform based methods~\cite{mono}. Multi pitch estimation is much harder as harmonic overtones are difficult to differentiate from the fundamental frequency~\cite{oud}. As the Fourier transform can be consistently performed within 1 millisecond on common hardware~\cite{fourierspeed}, it is well suited for real-time transcription.  % Because of this, most recent methods use machine learning or artificial intelligence. These methods are not well suited for real-time transcription. Also, little mathematical understanding of the problem is gained. Because of this, this paper will focus on mathematically solving this problem. Even if the methods researched in this paper might not outperform ML/AI solutions by themselves, they could be used to filter the input data for the ML/AI approaches. This might in turn improve the results of the AI/ML methods significantly.

Pitch estimation can be divided into a few subproblems. If a paper focuses on one of these subproblems, it also has to solve the other subproblems in order to test the system. Most papers do not publish their source code, which makes building upon other's work difficult. It also limits the research of the interaction between different solutions for the different subproblems. Because of this, we provide a framework in which can be built upon, as the implementation of a subproblem can easily be interchanged with another. The code can be found at "\url{github.com/lucmans/dechord}".

The goal of this paper is to give an overview of the problems when trying to solve this problem. This is done by developing a program which can translate a guitar signal to a set of observed notes in real-time. We combine techniques spread over different papers into one project and provide an algorithm which can estimate a set of played notes using the set of observed notes. The code should be publicly available and easily extensible so others can optimize a subproblem without having to solve the other subproblems. Using our program, we can experiment if Fourier transform based methods are feasible for real-time transcription.

Note that this paper is a preliminary paper for a possible master thesis. The focus is on building a prototype which can perform monophonic AMT and presenting a first draft of a polyphonic version. Using this information, we will estimate if the thesis project is feasible.

@@

A big problem in polyphonic pitch estimation is the occurrence of overtones~\cite{oud}. As these overtones also have a periodic nature in the frequency domain, they can be detected using another Fourier transform~\cite{double}. This does however have a big impact on the latency, as performing a Fourier transform is computationally expensive. Also, notes which are at octave intervals have the same periodicity~\cite{octave}, which makes them difficult to distinguish.

Another problem is choosing the real-time constraint for processing the input. Some papers claim 140 ms is sufficient~\cite{sloomboi}. This is however ridiculously long, as it would be about a note difference when playing eight notes in 200 BPM. Also, there already is a large latency due to how audio drivers work~\cite{os}. In practice, musicians often tweak their audio driver settings in order to gain 5 to 10 ms in latency, which makes the difference between being able to play faster pieces or not.

When limiting ourselves to transcribing a guitar, we can use physical limitations of the instrument to our advantage~\cite{physical}. However, this forces the guitarist to use the guitar in a specific way. For instance, these methods fail when the strings are tuned to different notes.

\section{Preliminaries}
In this paper, we use the Fourier transform to obtain a frequency domain representation of a signal. This transform takes a frame full of samples and produces a list of amplitudes corresponding to frequency bins. How this is done is described in Section~\ref{sub:four}.

When playing a note on an instrument, multiple frequencies are produced. The most notable is the fundamental frequency. Every note corresponds to one fundamental frequency. For instance, the fundamental frequency of the $\text{A}_4$ is 440 Hz. Integers multiples of the fundamental frequency resonate and produce harmonic overtones~\cite{overtones}. They are called harmonic overtones, as they are exact integer multiples and do not produce destructive interference among each other. Every integer multiple resonates with the fundamental frequency, however, each subsequent overtone is of lower amplitude.

Instruments do not produce perfect sine waves. The specific distortion of an instrument is called its timbre~\cite{timbre}. These distortions also show in the frequency domain. These are however very small compared to the fundamental frequency and its overtones of a note and can be disregarded as noise.

\section{Pitch estimation}
Automatic music transcription can be separated into two main problems; onset detection and pitch estimation~\cite{survey2}. In this paper, we will focus on pitch estimation. Because we always output the set of played notes for the last analysed frame, we do not require onset information. However, this information can improve the results of pitch estimation. For example, when an onset is detected, we can discard all samples from the transient.

As stated before, our goal is to perform the transcription in real-time. We choose a constraint of 6 ms. Most musicians work with a driver latency of 9 to 11 ms. On a good computer, this latency can be reduced to 3-5 ms. We will use these 6 ms of latency difference to perform our calculations in. Note that when playing more notes per second, latency becomes more of a problem. On slower pieces, we can get away with much larger latencies. In order to test the impact of extra latency for yourself, we developed a tool which plays back the incoming audio signal with a user specifiable delay. This tool can be found under the "latency" branch in our GitHub repository. Note that this tool adds a latency on top of the audio driver latency like our program would. The total latency is much larger than the displayed latency.

@@

\subsection{Transform to frequency domain}
To estimate the frequency components of a frame, transforms are used. Most commonly, the Fourier transform is used as it is the most researched transform and much is known about it~\cite{survey2}. Its main downside is that the frequency bins are constant in size, where the distance between notes increases exponentially. Because of this, low notes are hard to discern in the frequency domain where high notes have more resolution than needed. Because of this, other transforms such as the constant Q transform (CQT) are getting more popular.

When using transforms, window functions have to by applied to the frames to prevent spectral leakage~\cite{window}. Furthermore, the actual peak locations can be accurately interpolated using (Lagrangian) quadratic interpolation~\cite{interpol}. These concepts are further described in Section~\ref{sub:four} and Section~\ref{sub:peak} respectively.

\subsection{Peak picking}
Peaks in the frequency domain correspond to predominant frequencies in the analysed frame. However, not every peak is of interest. Some peaks are generated by noise. Many of these peaks can be filtered by requiring a minimal signal level to be considered a peak. Furthermore, due to spectral leakage, peaks can be observed which are not in the original signal. This kind of noise can be minimized by choose a good window function~\cite{window}. In polyphonic signals, harmonic overtones of different notes can interfere, which will lead to inconsistent and oscillating peaks.  % TODO: Maybe make preliminary section about overtone dissonance

On order to determine if a peak is significant enough, we first calculate a Gaussian average envelope~\cite{gauss}. For every point in a frame, the Gaussian average is a weighted average of all points in the frame. The weight is determined by the distance from the current point. The distance in number of bins goes through a Gaussian function to determine the actual weight. Here, higher values for $\sigma$ make points close by weight more compared to distant points. How this envelope is precisely calculated and used for peak picking is described in Section~\ref{sub:peak}.

\subsection{Note sets}
Given the spectral peaks from the previous step, we can try to calculate what notes they represent. Unfortunately, the found peaks do not only contain fundamental frequencies, but also harmonic overtones. This is only a small problem in monophonic transcription, as there is only one fundamental frequency and its overtones. However, in the polyphonic case, it is difficult to determine which peaks are fundamentals and by extension know how many notes are played at the same time.

To determine if notes are fundamentals or overtones, we use a "harmonic overtone sieve". Here, we look if any higher peak can be explained by the currently considered peak. The more other peaks a peak can explain, the more likely it is to be a fundamental. Moreover, if we cannot explain a certain peak, it is likely that there is a second note in the source signal.

\section{Implementation}  \label{sec:impl}
The program developed for this paper is written in C++ (11) and compiled with g++ (10.2.0). It uses SDL (2.0.14) for audio input/output and spectrum visualisation. FFTW (3.3.9) is used for efficient Fourier transforms. Its code can be found at "\url{github.com/lucmans/dechord}".

There are many parameters which change the behaviour of the overall system. For this paper, we choose some values which seemed to perform well. The values can still be optimized for better results. All parameters can be found in \texttt{config.h} along with comments explaining them.

By default, the program assumes an $A_4$ of 440 Hz. This can also be edited in \texttt{config.h}. Only 12 tone equal temperament (12-TET) tuning is currently supported. It could easily be changed to n-TET tuning, but non TET tuning is not easily supported.

An pseudo-code overview of the transcription process is given in Algorithm~\ref{alg:overview}.
\input{overview.pc}

\subsection{Transform to frequency domain}  \label{sub:four}
Our program uses the Fourier transform to obtain the frequency domain representation. We choose this transform as FFTW provides a very fast implementation. This speed is necessary for the real-time constrains on our system.

The resolution of the Fourier transform is dependent on the number of samples in a frame. We sample the guitar signal at 192 kHz, which is the maximum sample rate for most audio interfaces. To allow guitarists to play mildly fast, we have to limit our frame length. If the frame size if too large, multiple separate notes may be included in one transform, which would make it seem like the notes were played at the same time. Moreover, frame sizes which are a power of two can be transformed more efficiently, which further restricts are frame size choice. We chose a frame size of $2^{14} = 16384$ samples, which at 192 kHz is 85.33 ms (11.72 frames per second). This gives us a Fourier bin size of 11.72 Hz. At first glance, such a large bin size seems like a big problem, as the smallest interval that can be played on a guitar (G\#\textsubscript{2}-A\textsubscript{2}) is about 6 Hz. However, the overtones of these notes will have more Hz between them, which makes to possible to differentiate the two notes.% Also, as discussed in Section~\ref{sec:future}, we might be able to increase the frame size without reducing our frame rate too much.

Before applying the transform, a window function is often applied to minimize the artifacts produced by the transform. In this paper, we mainly use the Blackman-Nuttall window, but the following windows are also provided: Hamming window, Hann window, Blackman window, Nuttall window, Blackman-Harris window, Flat top window. Other window functions can easily be added.

Because our input data is purely real, we know the output of the Fourier transform will be symmetric. For this reason, we can omit calculation half of the output. This results in an almost two times speed-up\cite{twotimes}.

\subsection{Peak picking} \label{sub:peak}
Three different peak picking algorithms have been implemented. The first picks every local maximum (every \texttt{i} where \texttt{data[i - 1] < data[i] \&\& data[i] > data[i + 1]}). This approach finds too many peaks, especially in noise. The other two algorithms pick peaks based on a Gaussian average envelope. One algorithm picks all peaks which are above this average and the other only picks the highest peak for each spectral lobe above this average. Only picking the highest peak seemed to restrictive in testing, however, it is better resilient against noise and may be used with more careful parameter optimisation. By using the envelope, we effectively filter small local maxima close to large local maxima.

The Gaussian average envelope is calculated for every Fourier bin $i$ by taking the weighted average of a arbitrary number of bins around $i$. The weights are determined by a Gaussian function ($e^{-\pi(\frac{x}{\sigma})^2}$). The value of $\sigma$ and number of bins around $i$ can be changed in \texttt{config.h}. Higher values for $\sigma$ makes points around the considered point have a higher weighting. In our implementation, $\sigma$ is set to 1.2 and the number of bins is set to 47. In Algorithm~\ref{alg:gauss} and~\ref{alg:env}, a pseudo-code implementation is given for calculating the Gaussian weights and calculating the envelope based on these weights.
\input{gauss.pc}
\input{env.pc}

Even though this envelope works well for parts of the spectrum with peaks, it will still find peaks in noisy areas without peaks. This can be prevented in two ways. The first is setting a threshold which a peak has to exceed. This will result in less sustain, but works well. Secondly, we know that peaks outside the frequency range of a guitar cannot be a fundamental frequency. We can discard peaks lower than the lowest note, but peaks higher than the highest note still provide us information as it might be an overtone.

It is possible to determine the actual peak location within a Fourier bin using quadratic interpolation~\cite{interpol2}. Here, we calculate a value $p \in [-\frac{1}{2}, \frac{1}{2}]$, which is the location of the actual peak relative to the found peak in number of Fourier bins. In order to calculate $p$, we need the height of our found peak ($y(0)$) and the height of the two neighbouring bins ($y(-1)$ and $y(1)$). As quadratic interpolation is more accurate in the dB (logarithmic) scale, we define:
\begin{center}\vspace{-1mm}
    $\alpha = \log y(-1)$\\
    $\beta = \log y(0)$\\
    $\gamma = \log y(1)$
\end{center}\vspace{-1mm}
Then, we can calculate $p$ as follows:
\begin{center}\vspace{-1mm}
    $p = \frac{1}{2} \cdot \frac{\alpha - \gamma}{\alpha - 2\beta + \gamma}$
\end{center}\vspace{-1mm}
Given the Fourier bin size, we can calculate the frequency of the peak $i$ by multiplying the bin size by $i + p$.

\subsection{Note sets}  \label{sub:impnoteset}
Using all the found peaks, we can determine which peaks are overtones and which are fundamentals. First, we determine which notes the frequencies represent and calculate the error on these notes. The error is calculated as the number of cents from the closest harmonic note. The errors are different for every overtone (the series of errors is constant for every harmonic overtone series), as overtones are dissonant compared to equal temperament tuning. See Table~\ref{tab:overseries} for an example.
\begin{table}[H]\vspace{-3mm}
    \centering
    \begin{tabular}{rrcrr}
        $n$ & $f_{\text{overtone}}$ & closest note & $f_\text{note}$ & error \\
        \hline
        0 & 261.626  & $C_4$    & 261.626  &  0 \\
        1 & 523.251  & $C_5$    & 523.251  &  0 \\
        2 & 784.877  & $G_5$    & 783.991  &  1.955 \\
        3 & 1046.502 & $C_6$    & 1046.502 &  0 \\
        4 & 1308.128 & $E_6$    & 1318.510 &  -13.686 \\
        5 & 1569.753 & $G_6$    & 1567.982 &  1.955 \\
        6 & 1831.379 & $A^\#_6$ & 1864.655 &  -31.174 \\
    \end{tabular}
    \caption{Example of overtone series and the errors compared to the closest note}
    \label{tab:overseries}
\end{table}\vspace{-3mm}
If only harmonic notes were played on a guitar, we could filter all notes with an error higher than an arbitrary threshold. However, as most guitars are not perfectly intonated and strings detune by pressing them down on a fret, this information has to be used with care. Correctly using this information is out of the scope of this paper.

Creating a monophonic algorithm is relatively easy given the frequencies of the peaks. For instance, we can iterate over all found peaks and check if every higher peak can be a overtone of the considered peak. After counting the number of overtones for every peak, we can pick the peak with the most overtones.

Due to small errors, the overtones are not exactly integer multiples of the fundamental. As the distance between notes increases with frequency, we cannot set a constant threshold of error in Hz. Instead, we calculate the error in cents and choose a threshold for this error (configurable in \texttt{config.h}). This also allows for equal error tolerance above and below the note, which percentage based methods would not.

Polyphonic note detection can use a algorithm similar to the monophonic algorithm. As the number of played notes in unknown in the polyphonic case, we have to determine if a note is a fundamental or overtone. We cannot just set a threshold for the number of overtones a fundamental should have, because the number of detected overtones is very inconsistent. To solve this, we present a harmonic overtone sieve algorithm. Instead of directly looking for fundamentals, we try to sieve away all harmonics until we are left with fundamentals. We start at the lowest peak end mark all its possible overtones. If there are still unmarked peaks left, they are likely from another note and we sieve its overtones away. We can repeat this process until no more peaks can be sieved. If two peaks are not distinguishable in the root notes but are in the overtones, this algorithm will report the first distinguishable overtone as the root note. It is possible to calculate what the root note should have been, but that is outside the scope of this paper and will be further discussed in Section~\ref{sec:future}.

After this, we can remove all peaks which are outside of the frequency range of a guitar. The minimum and maximum can be changed in \texttt{config.h} to accommodate for different tunings of guitars with extra strings.

>>>>

% TODO
% Goal+contributie in abstract (ook real-time constaint)
% Contributies einde van intro

% Vragen aan Erwin
%        - Algemeen: Ik zie vaak dat je een uitleg verwacht eerder dan wanneer ik het zelf zou verwachten. In de intro/related work geef ik alleen een overview, later defineer je echt alles. Als je dit niet doet wordt je intro zo lang en technisch, wat mogelijk lezers afschrikt.
% Intro:
%        - Eerste alinea verwoord, beter zo?
%        - "is well suited for real-time transcrption". De intro lijkt mij niet de plek om hier heel uitgebreid op in te gaan, dat gebeurd ook later in de paper.
%        - "Most papers do not publish their source code, which makes building upon other's work difficult". Als een paper de methode "mathematically defined", betekend het niet dat het mogelijk is om erop verder te bouwen. De meeste papers zijn erg slordig in hun wiskunde noteren en het is ook niet altijd voor iedereen te begrijpen. Vaak missen er ook belangrijke details om iets net zo goed na te maken. Maar dit zijn slechts mogelijke verklaringen, in het veld zie ik verrassend weinig mensen op elkaars werk doorbouwen en zie ik veel mensen het wiel opnieuw uitvinden, wat duidelijk een probleem aantoont.
%        - Wat bedoel je precies met de real-time in quotes?
% Related work:
%        - Zijn de statements aan het einde van "related work" beter zo? (minder ad hoc)?
% Preliminaries:
%        - "You describe it as if this always happens...". Dat is toch ook zo?
% 5.1: Programming choises in different section. Dan wordt dit een sectie van 2 regels. Is het dan niet handiger om het gewoon in de corresponderende subsectie van implentation te laten?

% Antwoorden op jouw vragen
%       "experiment if Fourier methods are feasible for real-time transcription"->"is dat niet al gedaan?". Nee, in ieder geval niet praktisch. Er zijn een hoop papers die er een mening over hebben, maar weinig echte polyphone AMT systemen.
%       "accurately interpolate"->"estimate?". We interpoleren echt tussen 3 frequency bins. De hoogte van de frequency bins zouden eerder estimates zijn.
%       "such as the constant Q tranform"->"Why not use this tranform then?". Dit beschrijf ik als het goed is in de paper, maar omdat de Fourier erg snel is en veel onderzocht is.
%       "can be detected using another Fourier transform"->"what do you mean? explain". Voor nu iets anders verwoord, maar is het idee niet dat als je daar meer over wilt weten, je naar de referentie gaat?
%       Er missen bij veel referenties journal verwijzingen. In deze gevallen is het niet te vinden. Ik weet hier ook niet of het gewoon direct door een universiteit is uitgebracht of wat.
% 6.1: "Can the recordings not be improved?". Dat is een beetje vals spelen. Als ik dat zou doen, zou ik gewoon net zo lang kunnen blijven recorden totdat ik op zo'n manier speel dat het systeem 100% haalt.

@@

\section*{Abstract}
In this paper, we explore real-time transcription of an electric guitar signal. We focus on Fourier transform based methods, as this transform can be performed very fast. Using the obtained frequency domain, we select significant peaks using a Gaussian envelope. These peaks contain many overtones, which are filtered using our overtone sieve algorithm.

@@

% benadrukken dat er weinig nieuws is, maar combinatie is van bestaande dingen
\section{Introduction}
% Automatic Music Transcription (AMT) is a field of study which tries to automatically convert acoustic recordings into some form of (digital) music notation~\cite{survey1}. This task can be divided into two subproblems: pitch estimation and onset detection~\cite{survey2}. Instead of trying to convert a song into music notation, this paper will focus on translating a guitar signal into a set of played notes in real-time. This problem is very similar to AMT, as both require pitch estimation and onset/offset detection, however, we also have take latency into account.
Automatic Music Transcription (AMT) is a field of study which tries to automatically convert acoustic recordings into some form of (digital) music notation~\cite{survey1}. In this paper, we will focus on translating a guitar signal into a set of played notes in real-time. This task is similar to AMT, because both have to solve the problem of pitch estimation and onset/offset detection~\cite{survey2}. Because we try to solve the problem in real-time, we have some extra limitations. We must process the data fast enough such that no significant latency is introduced. Furthermore, we must process the data at least as fast as it is generated, otherwise latency would increase over time. Lastly, we cannot look-ahead, as can be done when transcribing pre-recorded songs.

Single pitch estimation can be accurately performed using Fourier transform based methods~\cite{mono}. Multi pitch estimation is much harder as harmonic overtones are difficult to differentiate from the fundamental frequency~\cite{oud}. As the Fourier transform can be performed consistently and efficiently on common hardware~\cite{fourierspeed}, it is well suited for real-time transcription.  % Because of this, most recent methods use machine learning or artificial intelligence. These methods are not well suited for real-time transcription. Also, little mathematical understanding of the problem is gained. Because of this, this paper will focus on mathematically solving this problem. Even if the methods researched in this paper might not outperform ML/AI solutions by themselves, they could be used to filter the input data for the ML/AI approaches. This might in turn improve the results of the AI/ML methods significantly.

Pitch estimation can be divided into a few subproblems. If a paper focuses on one of these subproblems, it also has to solve the other subproblems in order to test the system. Most papers do not publish their source code, which makes building upon other's work difficult. It also limits the research of the interaction between different solutions for the different subproblems. Because of this, we provide a framework which can be built upon, as the implementation of a subproblem can easily be interchanged with another. The code can be found at "\url{github.com/lucmans/dechord}".

The goal of this paper is to give a complete overview of the methods used to solve this problem. This is done by developing a program which can translate a guitar signal to a set of observed notes in real-time. We combine techniques spread over different papers into one project and provide an algorithm which can estimate a set of played notes using the set of observed notes. The code should be publicly available and easily extensible so others can optimize a subproblem without having to solve the other subproblems. Using our program, we can experiment if Fourier transform based methods are feasible for real-time transcription.

% Note that this paper is a preliminary paper for a possible master thesis. The focus is on building a prototype which can perform monophonic AMT and presenting a first draft of a polyphonic version. Using this information, we will estimate if the thesis project is feasible.
The focus of this paper is on building a prototype which can perform monophonic AMT and presenting a first draft of a polyphonic version.

@@

A big problem in polyphonic pitch estimation is the occurrence of overtones~\cite{oud}. As these overtones also have a periodic nature, they can be detected using another Fourier transform~\cite{double} on the obtained frequency domain. However, this periodicity also be detected with less overhead using note sets, which are explained in Section~\ref{sec:noteset_intro}. Notes which are at octave intervals have the same periodicity~\cite{octave}, which makes them difficult to distinguish. This is often referred to as the octave problem.

Another problem is choosing the real-time constraint for processing the input. Some papers claim 140 ms is sufficient~\cite{sloomboi}. This is however too long, considering it would be about a note difference when playing eight notes in 200 BPM. Furthermore, there always is a latency from the operating system due to how audio drivers work~\cite{os}.  % In practice, musicians often tweak their audio driver settings in order to gain 5 to 10 ms in latency, which makes the difference between being able to play faster pieces or not.

When focussing on transcribing a guitar, we can use physical limitations of the instrument to our advantage~\cite{physical}. However, by taking these limitation into consideration, the resulting program will be less versatile. For instance, these methods fail when the strings are tuned to different notes.

\section{Preliminaries}
In this paper, we use the Fourier transform to obtain a frequency domain representation of a signal. This transform takes a frame full of samples and produces a list of amplitudes corresponding to frequency bins. When performing the Fourier transform on a frame, spectral leakage is introduced~\cite{spectral}. This kind of noise is remedied by applying a window function to a frame. %How this is done is described in Section~\ref{sub:four}.

When playing a note on an instrument, multiple frequencies are produced. The most notable is the fundamental frequency. Every note corresponds to one fundamental frequency. Integer multiples of the fundamental frequency resonate and produce harmonic overtones~\cite{overtones}. They are called harmonic overtones, as they are exact integer multiples and do not produce destructive interference among each other. Every integer multiple resonates with the fundamental frequency, however, each subsequent overtone is of lower amplitude.  % For instance, the fundamental frequency of the $\text{A}_4$ is 440 Hz.

Generally, instruments do not produce perfect sine waves. The specific distortion of an instrument is called its timbre~\cite{timbre}. Even though the frequency components of these distortions are insignificant compared to the fundamental frequency and its overtones, it still results in extra noise in the signal. When strumming a note on the guitar, there is a strong attack. During this attack, the signal behaves chaotically and results in many fluctuating peaks in the frequency domain. This phenomenon is called a transient.

To specify in what octave a certain note lies, we use scientific note notation. This adds a subscript octave number to a note name. In scientific notation, we start from the note C. In other words, these notes are denoted in increasing pitch: $\text{C}_3$, $\text{A}_3$ and $\text{C}_4$. Note that $\text{A}_3$ is higher pitches than $\text{C}_3$, which is counter intuitive.

Most modern instruments use the twelve-tone equal temperament tuning system (12-TET). This tuning system divides an octave, which the distance between a frequency and its double, into twelve equal parts on the logarithmic scale. In other words, the ratio between two contiguous notes is $\sqrt[12]{2}$. The distance between two notes can further be divided on a logarithmic scale into 100 cents. The number of cents between two frequencies can be calculated as $1200 * \log_2(\frac{f_1}{f_2})$. 12-TET is tuned relative to a standard pitch. In modern music, we define the $\text{A}_4$ to be 440 Hz.

\section{Pitch estimation}
Automatic music transcription can be separated into two main problems; onset detection and pitch estimation~\cite{survey2}. In this paper, we will focus on pitch estimation. Onset detection is done implicitly, as we always output the set of played notes for the last analysed frame. If a note is added to the set of played notes, it can be considered as an onset. However, actual onset detection can improve the results of pitch estimation. For example, when an onset is detected, we can discard all samples from the transient that arises from this onset.

@@

\subsection{Transform to frequency domain}
To estimate the frequency components of a frame, transforms are used. Most commonly, the Fourier transform is used as it is the most researched transform and much is known about it~\cite{survey2}. Its main downside is that the frequency bins are constant in size, where the distance between notes increases exponentially. Because of this, low notes are hard to discern in the frequency domain where high notes have more resolution than needed. Because of this, other transforms such as the constant Q transform (CQT) are getting more popular. Despite this, we will still use the Fourier transform as very efficient implementations exist for this transform. We need this efficiency to remain real-time.

%When using transforms, window functions have to by applied to the frames to prevent spectral leakage~\cite{window}. Furthermore, the actual peak locations can be accurately interpolated using (Lagrangian) quadratic interpolation~\cite{interpol}.% These concepts are further described in Section~\ref{sub:four} and Section~\ref{sub:peak} respectively.
When using the Fourier transform, a window function has to be applied to a frame to prevent spectral leakage~\cite{window}. In this paper, we use the Blackman-Nuttall window. Given a window with $N$ samples, the Blackman-Nuttall window $w(n)$ is defined as:\vspace{-1mm}
\begin{center}
    \hspace{-12mm} $w(n) = 0.3635819 - 0.4891775*\cos(\frac{2\pi n}{N})$ \\\vspace{+0.5mm}
    \hspace{+4mm} $+ 0.1365995 * \cos(\frac{4\pi n}{N}) - 0.0106411 * \cos(\frac{6\pi n}{N})$
\end{center}%\vspace{-2.5mm}
Then, with the signal $s(n)$ and the window $w(n)$, we can obtain the resulting windowed signal res$(n)$ using:
\begin{center}%\vspace{-2mm}
    $\text{res}(n) = s(n) * w(n)$
\end{center}
Figure~\ref{fig:spec} shows the effect of a window function on a frame graphically.
\begin{figure}[H]%\vspace{-2mm}
    \centering
    \includegraphics[width=\linewidth]{3.png} %0.85
    \caption{The effect of applying a window function to a signal}
    \label{fig:spec}
\end{figure}%\vspace{-4mm}
%In polyphonic signals, harmonic overtones of different notes can interfere, which will lead to inconsistent and oscillating peaks.  % TODO: Maybe make preliminary section about overtone dissonance

The output of a Fourier transform is a list of complex number. The norm of each complex number corresponds to the amplitude of a certain frequency bin. The size of these frequency bins $f_{\text{bin}}$ is determined by the sampling rate $f_{\text{SR}}$ of the signal (samples per second) and the number of samples in a frame SPF with the following relation:
\begin{center}
    $f_{\text{bin}} = \frac{f_{\text{SR}}}{\text{SPF}}$
\end{center}
Furthermore, the frame length $t_\text{frame}$ in seconds can be calculated as follows:
\begin{center}
    $t_{\text{frame}} = \frac{\text{SPF}}{f_{\text{SR}}}$
\end{center}
% Higher sampling rates give us more precision, but also requires more processing power. Also, widespread audio interfaces only offer sampling rates up to 192 kHz.

\subsection{Peak picking}
Peaks in the frequency domain correspond to predominant frequencies in the analysed frame. A peak is defined as a bin which has a higher amplitude than its left and right neighbour.

Not every peak is of interest. Some peaks are generated by noise in the signal. Many of these peaks can be filtered by requiring an arbitrary signal level to be considered a peak. This approach will however result in less sustain.

In order to determine if a peak is significant enough, we first calculate a Gaussian average envelope~\cite{gauss}. For every frequency bin, the Gaussian average is a weighted average of all frequency bins, where the weight is determined by the distance from the considered bin. Given a distance of $n$ bins, the weight is determined by the following Gaussian function:
\begin{center}
    $e^{-\pi(\frac{n}{\sigma})^2}$
\end{center}
Here, $\sigma$ can arbitrarily be chosen. Higher values for $\sigma$ make points close by weight more compared to distant points. By using the envelope, we effectively filter small local maxima close to large local maxima.

Every peak which surpass the envelope and is above the arbitrary signal level is deemed significant and used in the next step, the harmonic overtone sieve. We also provided a different variant of the peak picker where only the highest peak is selected for every lobe which surpasses the envelope and a variant which picks all peaks, disregarding the envelope. We also provided a peak picker which picks every peak, which is useful for development purposes.

\subsection{Harmonic overtone sieve}  \label{sec:noteset_intro}
Given the significant spectral peaks from the previous step, we can try to calculate what notes they represent. First, we get the actual location of a peak with a Fourier bin using quadratic interpolation~\cite{interpol}\cite{interpol2}. Here, we calculate a value $p \in [-\frac{1}{2}, \frac{1}{2}]$, which is the location of the actual peak relative to the found peak in number of Fourier bins. In order to calculate $p$, we need the height of our found peak ($y(0)$) and the height of the two neighbouring bins ($y(-1)$ and $y(1)$). As quadratic interpolation is more accurate in the dB (logarithmic) scale, we define:
\begin{center}%\vspace{-3mm}
    $\alpha = \log y(-1)$\\
    $\beta = \log y(0)$\\
    $\gamma = \log y(1)$
\end{center}%\vspace{-1mm}
Then, we can calculate $p$ as follows:
\begin{center}%\vspace{-1mm}
    $p = \frac{1}{2} \cdot \frac{\alpha - \gamma}{\alpha - 2\beta + \gamma}$
\end{center}%\vspace{-1mm}
Given $p$, we can also calculate the amplitude $a$ corresponding to the interpolated location:
\begin{center}
    $a = \beta - \frac{(\alpha - \gamma) * p}{4}$
\end{center}
Now, we can calculate the frequency of the peak $i$ by multiplying the Fourier bin size by $i + p$. Using the obtained frequencies and corresponding amplitudes, we construct note sets. A note contains all useful information about the peak for transcription. Other than the interpolated frequency and corresponding amplitude, we calculate the distance to the closest note in cents and we also store the scientific pitch notation. To calculate the distance between our interpolated frequency $f_i$ and the closest note, we first calculate the tuned frequency $f_t$:
\begin{center}
    $f_t = f_{\text{A}_4} * 2^{\text{round}(12 * \log_2(\frac{f_i}{f_{\text{A}_4}})) / 12}$  %exp2(round(12.0 * l2) / 12.0)
\end{center}
Then, we can calculate the distance $d$:
\begin{center}
    $d = 1200 * \log_2(\frac{f_i}{f_t})$
\end{center}
Given an array with all the note names (A, A\#, B, C, C\# etc.), we can calculate the index which contains the name of the note as follows:
\begin{center}
    $i = \text{round}(12 * \log_2(\frac{f_i}{f_{\text{A}_4}}) \mod 12)$
\end{center}
Here, we assume that "$-1 \mod 12 = 11$", which is not the case in all programming languages. To calculate the octave number, we first calculate the pitch of $\text{C}_1$:
\begin{center}
    $f_{\text{C}_1} = f_{\text{A}_4} * 2^{\frac{45}{12}}$
\end{center}
Using this, we can calculate the octave number $o$:
\begin{center}
    $o = \text{floor}(\log_2(\frac{f_i}{f_{\text{C}_1}}) + 1)$
\end{center}
Note that this equation fails if we calculate the octave number of a $C$ which is tuned too low ($f_i < f_t$). To remedy this, we can simply add a check if this is the case and increment the octave number accordingly.
%
% Usually, peaks do not exactly coincide with a note. In order to calculate the error in cents, we need the tuned frequency $f_t$ of the closest note to our observed frequency $f_o$. First, we have to calculate the number of semitones $n$ ($\sqrt[12]{2}$ steps) it is removed from a tuning note, such as the $\text{A}_4$:
% \begin{center}
%     $n = \text{round}(12 * \log_2(\frac{f_o}{f_{\text{A}_4}}))$
% \end{center}
% The decimal part of $n$ before rounding would be the amount of dissonance. By rounding this number, we find the closest note. Now, we can calculate the tuned frequency by doing the inverse of the previous calculation:
% \begin{center}\vspace{-3mm}
%     $f_t = f_{\text{A}_4} * 2^{\frac{n}{12}}$
% \end{center}
% Using this tuned frequency, we can calculate the error $e$ in cents as follows:
% \begin{center}
%     $e = 1200 * \log_2(\frac{f_o}{f_t})$
% \end{center}
% If the guitar is tuned and intonated correctly, we can set a threshold on the error of a detected note. Only notes within this threshold can be fundamental notes.

Unfortunately, a note set does not only contain fundamental frequencies, but also harmonic overtones. This is only a small problem in monophonic transcription, as there is only one fundamental frequency and its overtones. However, in the polyphonic case, it is difficult to determine which peaks are fundamentals and by extension know how many notes are played at the same time.

To determine if notes are fundamentals or overtones, we developed a "harmonic overtone sieve" algorithm. This algorithm iterates over all found peaks and checks if any higher pitched peak is a possible overtone of the considered peak. As the calculated frequency of a peak may not be accurate, we have to set a threshold on the error under which a peak is considered an overtone. Because the distance between notes increases with frequency, we cannot set a constant threshold of error in Hz. Therefore, we calculate the error in cents and choose a threshold for this error. This also allows for equal error tolerance above and below the note, which percentage based methods would not. Given the frequency of the considered peak $f_c$ and the possible overtone $f_o$, we can calculate the frequency of the closest harmonic overtone $f_h$:
\begin{center}
    $f_h = f_c * \text{round}(\frac{f_o}{f_c})$
\end{center}
Using the closest harmonic overtone frequency, we can calculate the error from our possible overtone in cents:
\begin{center}
    $e = 1200 * \log_2(\frac{f_o}{f_h})$
\end{center}
T

For every peak, we keep track of two variables. The first is the number of number of overtones we detected for every peak. The second is a boolean which indicates if a peak is a overtone of another peak. For monophonic note detection, we can assume that the peak with the most overtones is the fundamental frequency. For polyphonic note detection, we can assume that every peak which is not an overtone of another lower pitches peak is a fundamental. Another method for polyphonic note detection would be setting a threshold for the number of overtones a fundamental should have. This however gives inconsistent results, because when the same note is played louder, more overtones are detected. Furthermore, the number of overtones decreases while a note sustains.

A guitar has a limited range of fundamental tones which it can produce. Using this information, we can discard all peaks lower than the lowest fundamental of a guitar as noise and mark all peaks higher than the highest fundamental as overtones or noise. These ranges have to be configurable to accommodate different tunings or guitars with extra strings or frets.
